Fail Faster

### 2025-03-23 paper/paper.tex
change recursive explanation

The type-deriving algorithm follows a compositional pattern. Generators for complex types 
are synthesized by structurally composing generators of their subtypes: 
base types are mapped to primitive generators included in the generator library;
product types---such as tuples and records---are handled by sampling each component 
using \texttt{bind} and aggregating the results; sum types, or variants, are generated using 
a \texttt{weighted\_union} of the generators for each case; and recursive types are handled by 
tying the knot at the generator level---recursive references are treated as parameters during 
derivation, allowing the generator's structure to be defined first and then closed using a 
fixed-point combinator.

This compositional approach maps naturally to staged generation: 
to derive a staged generator, we replace each standard combinator 
with its staged counterpart. The derivation algorithm remains unchanged.

Our implementation in OCaml uses the preprocessor extension---``PPX''---system, but the 
approach is portable to other languages with metaprogramming capabilities. 
In Scala, for instance, the same strategy can be realized using typeclass 
resolution. In both settings, the result is a three-stage process: a 
metaprogram---via PPX or typeclasses---constructs a generator expression 
composed of staged combinators; this expression is evaluated at compile 
time to produce a specialized generator; and finally, the generator is 
executed at runtime to produce values. The entire 
pipeline remains invisible to the user.

>>>>

The type-deriving algorithm follows a compositional pattern. Generators for complex types 
are synthesized by structurally composing generators of their subtypes: 
base types are mapped to primitive generators included in the generator library;
product types---such as tuples and records---are handled by sampling each component 
using \texttt{bind} and aggregating the results; sum types, or variants, are generated using 
a \texttt{weighted\_union} of the generators for each case; and for recursive types, 
the entire generator is wrapped in a fixed-point combinator, and a recursive handle 
is used as the component generator for all recursive occurrences of the type.
This compositional approach maps naturally to staged generation: 
to derive a staged generator, we replace each standard combinator 
with its staged counterpart. The derivation algorithm remains unchanged.

Our implementation in OCaml uses the preprocessor extension---``PPX''---system, but the 
approach is portable to other languages with metaprogramming capabilities. 
In Scala, for instance, the same strategy can be realized using typeclass 
resolution. In both settings, the result is a three-stage process: a 
metaprogram---via PPX or typeclasses---constructs a generator expression 
composed of staged combinators; this expression is evaluated at compile 
time to produce a specialized generator; and finally, the generator is 
executed at runtime to produce values. The entire 
pipeline remains invisible to the user. \tr{TODO: Boolean list example in a figure...}
